<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{labelled}
-->

# Introduction to labelled

The purpose of the **labelled** package is to provide functions to manipulate
metadata as variable labels, value labels and defined missing values
using the `labelled` class and the `label` attribute introduced in `haven` package.

## Variable labels

A variable label could be specified for any vector using `var_label`.

```{r}
library(labelled)

var_label(iris$Sepal.Length) <- "Length of sepal"
```

It's possible to add a variable label to several columns of a data frame using a named list.

```{r}
var_label(iris) <- list(Petal.Length = "Length of petal", Petal.Width = "Width of Petal")
```

To get the variable label, simply call `var_label`.

```{r}
var_label(iris$Petal.Width)
var_label(iris)
```

To remove a variable label, use `NULL`.

```{r}
var_label(iris$Sepal.Length) <- NULL
```

In **RStudio**, variable labels will be displayed in data viewer.

```{r, eval=FALSE}
View(iris)
```

## Value labels

The first way to create a labelled vector is to use the `labelled` function.
It's not mandatory to provide a label for each value observed in your vector.
You can also provide a label for values not observed.

```{r}
v <- labelled(c(1,2,2,2,3,9,1,3,2,NA), c(yes = 1, no = 3, "don't know" = 8, refused = 9))
v
```

Use `val_labels` to get all value labels and `val_label` to get the value label associated 
with a specific value.

```{r}
val_labels(v)
val_label(v, 8)
```

`val_labels` could also be used to modify all the value labels attached to a vector, 
while `val_label` will update only one specific value label.

```{r}
val_labels(v) <- c(yes = 1, nno = 3, bug = 5)
v
val_label(v, 3) <- "no"
v
```

With `val_label`, you can also add or remove specific value labels.

```{r}
val_label(v, 2) <- "maybe"
val_label(v, 5) <- NULL
v
```

To remove all value labels, use `val_labels` and `NULL`. The `labelled` class will also be removed.

```{r}
val_labels(v) <- NULL
v
```

Adding a value label to a non labelled vector will apply `labelled` class to it.

```{r}
val_label(v, 1) <- "yes"
v
```

Note that applying `val_labels` to a factor will have no effect!

```{r}
f <- factor(1:3)
f
val_labels(f) <- c(yes = 1, no = 3)
f
```

You could also apply value labels to several columns of a data frame.

```{r}
df <- data.frame(v1 = 1:3, v2 = c(2, 3, 1), v3 = 3:1)

val_label(df, 1) <- "yes"
val_label(df[, c("v1", "v3")], 2) <- "maybe"
val_label(df[, c("v2", "v3")], 3) <- "no"
val_labels(df)

val_labels(df[, c("v1", "v3")]) <- c(YES = 1, MAYBE = 2, NO = 3)
val_labels(df)
val_labels(df) <- NULL
val_labels(df)
val_labels(df) <- list(v1 = c(yes = 1, no = 3), v2 = c(a = 1, b = 2, c = 3))
val_labels(df)
```

## Sorting value labels

Value labels are sorted by default in the order they have been created.

```{r}
v <- c(1,2,2,2,3,9,1,3,2,NA)
val_label(v, 1) <- "yes"
val_label(v, 3) <- "no"
val_label(v, 9) <- "refused"
val_label(v, 2) <- "maybe"
val_label(v, 8) <- "don't know"
v
```

It could be useful to reorder the value labels according to their attached values.

```{r}
sort_val_labels(v)
sort_val_labels(v, decreasing = TRUE)
```

If you prefer, you can also sort them according to the labels.

```{r}
sort_val_labels(v, according_to = "l")
```

## Converting to NA

In some cases, values who don't have an attached value label could be considered as missing.
`nolabel_to_na` will convert them to `NA`.

```{r}
v <- labelled(c(1,2,2,2,3,9,1,3,2,NA), c(yes = 1, maybe = 2, no = 3))
v
nolabel_to_na(v)
```

In other cases, a value label is attached only to specific values that corresponds to 
a missing value. For example:

```{r}
size <- labelled(c(1.88, 1.62, 1.78, 99, 1.91), c("not measured" = 99))
size
```

In such cases, `val_labels_to_na` could be appropriate.

```{r}
val_labels_to_na(size)
```

These two functions could also be applied to an overall data frame. 
Only labelled vectors will be impacted.

## Converting to factor

A labelled vector could easily be converted to a factor with `to_factor`.

```{r}
v <- labelled(c(1,2,2,2,3,9,1,3,2,NA), c(yes = 1, no = 3, "don't know" = 8, refused = 9))
v
to_factor(v)
```

The `levels` argument allows to specify what should be used as the factor levels, 
i.e. the labels (default), the values or the labels prefixed with values.

```{r}
to_factor(v, levels = "v")
to_factor(v, levels = "p")
```

The `ordered` argument will create an ordinal factor.

```{r}
to_factor(v, ordered = TRUE)
```

The argument `nolabel_to_na` specify if the corresponding function should be 
applied before converting to a factor. Therefore, the two following commands are equivalent.

```{r}
to_factor(v, nolabel_to_na = TRUE)
to_factor(nolabel_to_na(v))
```

`sort_levels` specifies how the levels should be sorted: `"none"` to keep the order in which value labels 
have been defined, `"values"` to order the levels according to the values and `"labels"` according 
to the labels. `"auto"` (default) will be equivalent to `"none"` except if some values with no 
attached labels are found and are not dropped. In that case, `"values"` will be used.

```{r}
to_factor(v, sort_levels = "n")
to_factor(v, sort_levels = "v")
to_factor(v, sort_levels = "l")
```

The function `to_labelled` could be used to turn a factor into a labelled numeric vector. 

```{r}
f <- factor(1:3, labels = c("a", "b", "c"))
to_labelled(f)
```

Note that `to_labelled(to_factor(v))` will not be equal to `v` due to the way factors are
stored internally by **R**.

```{r}
v
to_labelled(to_factor(v))
```

## Importing labelled data

In **haven** package, `read_spss`, `read_stata` and `read_sas` are natively importing data
using the `labelled` class and the `label` attribute for variable labels.

Functions from **foreign** package could also import some metadata from **SPSS** and **Stata**
files. `to_labelled` can convert data imported with **foreign** into a labelled data frame.
However, there are some limitations compared to using **haven**:

- For **SPSS** files, it will be better to set `use.value.labels = FALSE`, `to.data.frame = FALSE` and
  `use.missings = FALSE` when calling `read.spss`. If `use.value.labels = TRUE`, variable with
  value labels will be converted into factors by `read.spss` (and kept as factors by `foreign_to_label`). 
  If `to.data.frame = TRUE`, meta data describing the missing values will not be imported. 
  If `use.missings = TRUE`, missing values would have been converted to `NA` by `read.spss`.
- For **Stata** files, set `convert.factors = FALSE` when calling `read.dta` to avoid conversion
  of variables with value labels into factors. So far, missing values defined in Stata are always 
  imported as `NA` by `read.dta` and could not be retrieved by `foreign_to_labelled`.
  
The **memisc** package provide functions to import variable metadata and store them in specific
object of class `data.set`. The `to_labelled` method can convert a data.set into a labelled data frame.

```{r, eval=FALSE}
  # from foreign
  library(foreign)
  df <- to_labelled(read.spss(
    "file.sav",
    to.data.frame = FALSE,
    use.value.labels = FALSE,
    use.missings = FALSE
 ))
 df <- to_labelled(read.dta(
   "file.dta",
   convert.factors = FALSE
 ))

 # from memisc
 library(memisc)
 nes1948.por <- UnZip("anes/NES1948.ZIP", "NES1948.POR", package="memisc")
 nes1948 <- spss.portable.file(nes1948.por)
 df <- to_labelled(nes1948)
 ds <- as.data.set(nes19480)
 df <- to_labelled(ds)
```

