---
author: "Joseph Larmarange"
title: "Generate a data dictionnary and search for variables with `look_for()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate a data dictionnary and search for variables with `look_for()`}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Showing a summary of a data frame

### Default printing of tibbles

It is a common need to easily get a description of all variables in a data frame.

When a data frame is converted into a tibble (e.g. with `dplyr::as_tibble()`), it as a nice printing showing the first rows of the data frame as well as the type of column.

```{r message=FALSE}
library(dplyr)
```


```{r}
iris %>% as_tibble()
```

However, when you have too many variables, all of them cannot be printed and their are just listed.

```{r}
data(fertility, package = "questionr")
women
```

Note: in **R** console, value labels (if defined) are usually printed but they do not appear in a R markdown document like this vignette.

### `dplyr::glimpse()`
 
 The function `dplyr::glimpse()` allows you to have a quick look at all the variables in a data frame.
 
```{r}
glimpse(iris)
glimpse(women)
```
 
It will show you the first values of each variable as well as the type of each variable. However, some important informations are not displayed:

- variable labels, when defined;
- value labels for labelled vectors;
- the list of levels for factors;
- the range of values for numerical variables.

### `labelled::look_for()`

```{r}
library(labelled)
look_for(iris)
look_for(women)
```


